Manuscript title here

### 2024-09-23 project-plan.Rmd
uppdaterad metod enligt manuscript

To differentiate the severity of trauma shock, we can divide shock into different degrees. One of such classification is the ATLS classification for major haemorrhage. The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However, this only meant to be used for hypovolemic shocks and not the other types. There is other classification for predicting massive transfusion (TASH, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2). The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock. ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).

One way to improve trauma care quality is trough identifying Opportunities for Improvement (OFI). This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals. There are different processes to finding OFIs, but one of the primary ways is structured multidisciplinary morbidity and mortality reviews of patient cases [@world_health_organization_guidelines_2009]. In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988]. It is based on the three factors of structure, process and outcome. By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?]. This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].

@@

*Refer to the appropriate reporting guideline for details. If you are
developing, updating or validating a clinical prediction model then use
[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/).
If you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*

Study design We conducted a registry based retrospective cohort study,
using data from the trauma registry and trauma care quality database at
the Karolinska University Hospital in Solna. We then linked the two
databases to assess what types of OFI occurred in patients arriving in
shock and use adjusted logistic regression to assess the association
between the degree of shock and odds of opportunities for improvement.

Setting The trauma registry includes about 14000 patients treated at
Karolinska University Hospital in Solna between 2012 and 2023. The
trauma care quality database is a subset of the trauma registry and
includes about 8000 patients selected for review between 2014 and 2023.
Karolinska University Hospital in Solna acts as a level one trauma
center according to critera set by American College of Surgeons(), and
treats all major trauma in the greater metropolitan area of Stockholm.

Participatns Every trauma patient at Karolinska University Hospital is
included in a morbidity and mortality review process, which involves
both individual case evaluations by specialized nurses and audit
filters. Patients identified with a high potential for OFIs are
discussed at multidisciplinary conferences. The identified OFIs are then
categorized into broader areas. The multidisciplinary conferences are
held every six to eight weeks, during which an average of ten patient
cases are reviewed by experienced specialists from all trauma-related
fields. The presence or absence of OFIs is determined by consensus among
all participants and is documented in the trauma care quality database.

To qualify into the trauma registry, requires admission by trauma team
activation or any admission with an Injury Severity Score (ISS) of more
than nine. We included all patients in ... between ... and ... We
excluded patients younger than 18 and/or were dead on arrival.

Variable and measurements To classify the patients "heart rate", "blood
pressure", "respiratory rate" and "mental status" will be used as
independent variables for classification of shock regardless of their
cause. Based on these four measurements, the patients will be classified
into four groups roughly based on the same value as in the ATLS trauma
shock classifications. The outcome will be OFI recorded in the trauma
care quality database and are binary when used to calculate the
association with degree of shock. The subgroups of OFI will be explored
and reported in a list, and therefore non-binary.

>>>>

To differentiate the severity of trauma shock, we can divide shock into different degrees. One of such classification is the ATLS classification for major haemorrhage. The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However, this only meant to be used for hypovolemic shocks and not the other types. There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2). The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock. ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).

One way to improve trauma care quality is trough identifying Opportunities for Improvement (OFI). This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals. There are different processes to finding OFIs, but one of the primary ways is through structured multidisciplinary morbidity and mortality reviews of patient cases [@world_health_organization_guidelines_2009]. In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988]. It is based on the three factors of structure, process and outcome. By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?]. This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].

@@

If you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*

## Study design

<!-- Present key elements of study design -->

We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.

## Setting

The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2014 and 2023. The "problem" database is a subset of the trauma registry and includes about ? patients selected for review. The Karolinska University Hospital in Solna treats all major trauma in the greater metropolitan area of Stockholm.

## Participants

To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.

Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters. Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences. The identified OFIs are then categorized into broader areas. The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields. The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.

We included all patients in "SWETRAU" and "Problem" between 2014 and 2023 We excluded patients younger than 15 and/or were dead on arrival.

## Variables and data sources/measurements

<!-- Elaborate on the outcome OFI. How is it defined? What are the different types of OFI? -->

The outcome was defined as presence of atleast one OFI, determined by the multidisciplinary M&M conference in the "problem" registry. An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, "preventable death" (den här var väl inte ens med i "problem", fast fanns med i studien som Martin delade på discord) and other errors.

In this study, the outcomes will be studied in two ways, one is the amount of occurrence for various types of OFI in shock patients, and the other is binary where odds of OFI for every shock class i calculated out of presence of OFI or not. (kanske inte ens en del av metod)

To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause. Heart rate(HR), systolic blood pressure(SBP), respiratory rate(RR) and mental status. Based on these four measurements, the patients will be classified into four groups roughly based on the same value as in the ATLS trauma shock classifications.

<!-- Elaborate on how you will categorise the degree of shock, what are the different levels?-->

Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].

Version 1, 
Due to the tenth edition of ATLS classifications lack actual numbers, we will be using simplified and numerical version approximated by Dunham et al[@dunham_comparison_2017]. (INSERT PICTURE)

Version 2, 
Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values. For HR and RR we will be using the numbers from the ninth edition of the classification. The SBP will be divded into class I and II where the SBP is >110, class III 110-90mmhg and class IV <90mmhg. This selection is based on tenth edition, where SBP is defined as normal i class I to II and normal/below in class III and clearly below normal i class IV. These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90mmgh and class IV below 90mmhg. For mental status, according to the tenth edition ATLS, class I and II should have normal GCS, and class III and IV have lower GCS. Therefore we decided to use GCS 15 for class I + II and GCS <15 for class III and IV.

 - Hur gör vi med de som hamnar mellan klasserna?
To classify, all of the parameters in one class needs to be fulfilled. In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017]. At same time ATLS manual states one of the common pitfalls is "Diagnosis of shock can be missed when only a single parameter is used." therefore to improve sensitivity we decided to use the worst parameter for classification.

To prevent classifying non-shock traumatic brain injury as class III and worse shock based on GCS, we decided to only include those who also present a circulatory parameter in accordance with class III or IV. 
 - eller ta bort GCS helt och hållet? För i enlighet med ovan kommer GCS alltid att prioriteras bort och då finns ingen anlending att ha med det...

Version 3, simplaste och den jag gillar mest, men kan andra sjukdomar som gluykolen förgiftning sabba tanken?
We will be only using the base deficit parameter according to the tenth edition of the ATLS classification for haemorrhage. There are three reasons. The first one is ATLS tenth edition lack numerical values for vital parameters.  The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017]. The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause. The classes will be divided into class I: 0 - (-2), class II: (-2)-(-6), class III: (-6)-(-10), and class IV: <(-10).

## Bias

## Study size

All available data in "SWETRAU" and "Problem" database will be included.

## Quantitative variables

## Statistical methods

The statistical analysis will be performed using R, a programing language and environment for statistical computing. We will present the type of OFI as percentage distributions and then visualize it in a pie chart. Bivariable logistic regression will be used to determine the association between the OFI and degree/class of shock according to ATLS. It will be presented as odds ratios (OR) between presence of OFI, and shock classes. The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used. All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the “problem” database to ensure objectivity.
